---
title: "Working with VCF files"
author: "Jose V. Die"
output: html_notebook
---
<br> 
  
Quick look at the file  
```{r}
cat chr20.vcf | less
```
&nbsp;  

Get the columns header number of lines 
```{r}
grep -n CHROM chr20.vcf
```
&nbsp; 

Get first data (from that number of line)
```{r}
sed -n 67,100p chr20.vcf | less -S
```
&nbsp; 

**Question: How many variants are found in Chr20?**
Number of total lines in the file:
```{r}
wc -l chr20.vcf
```
 
Number of variants: number of lines from the 1st variant to the last one. 
```{r}
sed -n 68,152562p chr20.vcf | wc -l
```

Extract from the header the filter descriptions
```{r}
grep FILTER chr20.vcf | less -S
```
&nbsp; 

**Question: How many of these variants have a “FILTER” value of “PASS”?**
We know that the Filter filed is column 7 in a VCF file, so:  
```{r}
sed -n 68,152562p chr20.vcf | cut -f 7 | grep -c PASS
```
&nbsp; 

**Question: How many of the variants with a “FILTER” value of “PASS” are phased in NA12877?**
(Hint:Count only the SNVs with a "|" symbol in their genotype.). 

```{r}
sed -n 67,152562p chr20.vcf  | cut -f 7,10 | grep PASS | cut -c -1,-8 | grep "|" | wc -l
```
&nbsp; 

Databases that store information about variants use unique identifiers, called “rsIDs”. 
Using rsIDs instead of chromosomal position to identify SNVs allows for the identification of the SNVs even as the reference genome is updated and chromosomal positions change. 
**Question: How many SNVs with a “FILTER” value of “PASS” have an rsID?**
```{r}
sed -n 67,152562p chr20.vcf  | cut -f 1,-7 | grep PASS | grep "rs" | head
sed -n 67,152562p chr20.vcf  | cut -f 1,-7 | grep PASS | grep "rs" | wc -l
```
&nbsp; 

Alagille syndrome is a rare genetic disorder characterized by liver damage caused by abnormalities in the bile ducts. 
It is caused by a mutation in the JAG1 gene, which is located on chromosome 20. Recall that we used the GRCh38 reference 
genome to perform variant calling, and with respect to GRCh38, the JAG1 gene starts at position 10,637,684 and ends at 
position 10,674,046 of chromosome 20.

**Question: How many variants with a “FILTER” value of “PASS” does NA12877 have in the JAG1 gene?**
First, we have to figure out the lines containing the gene JAG1. There are several ways; here just an alternative:
```{r}
# first line containing the gene
grep 106379 chr20.vcf 
grep -n 10637926 chr20.vcf 
# last line containing the gene
 grep 106739 chr20.vcf 
grep -n 10673963 chr20.vcf 
```

```{r}
sed -n 26852,26970p chr20.vcf | grep PASS | cut -f1,-7 | wc -l
```

If an indel in a coding region is not a multiple of three nucleotides in length, the resulting mRNA will be out-of-frame beginning at the indel, resulting in a **frameshift mutation**. 
If an indel or mutation changes an amino-acid-coding codon into a STOP codon resulting in a **nonsense mutation**, the resulting protein will be truncated at the indel site. 
**Question: How many variants with a “FILTER” value of “PASS” are frameshift variants (denoted by “frameshift_variant” in the VCF file)?**
```{r}
grep PASS  chr20.vcf | grep frameshift_variant | wc -l
```
**Question: How many are nonsense variants (denoted by “stop_gained” in the VCF file)?**
```{r}
grep PASS  chr20.vcf | grep stop_gained | wc -l
```


  
